Interisland movements of woodland caribou in Newfoundland

### 2020-02-19 paper/manuscript.Rmd
Update manuscript.Rmd and manuscript.docx

---
title: Interisland movements of woodland caribou in Newfoundland
author: Quinn M.R. Webber^1\*^, Jack G. Hendrix^1^, Alec L. Robitaille^2^,  Eric Vander Wal^1,2^
date: |
  | ^1^Cognitive and Behavioural Ecology Interdisciplinary Program, Memorial University of Newfoundland
  | ^2^Department of Biology, Memorial University of Newfoundland, Canada
  | \*Corresponding author: Quinn M.R. Webber (webber.quinn@gmail.com)
output: 
    github_document
  # bookdown::pdf_document2:
  #    number_sections: true
  #    toc: false
  #bookdown::word_document2:
  #  reference_docx: template.docx
bibliography: ../references/knitcite.bib
#csl: csl.csl
---

@@

Caribou (*Rangifer tarandus*) are exceptional swimmers. Ample evidence exists
that caribou swim in streams, rivers, and lakes during migration (Leblond et al.
2016) to avoid predators [@Bergerud_1985] and access islands during calving
(Bergerud et al. 1990). Swimming for quadrupedal terrestrial mammals can be
thought of as a set of trade-offs. Costs include high energy expenditure and
risk of drowning. Specifically, even though caribou are adept swimmers, the
energetic expenditure associated with swimming for quadrupedal mammals is
significantly higher than walking or running (Fish 1993), whiledrowning
is also possible (Miller & Gunn 1985). Despite a large number of coastal and
island caribou herds, only a handful of examples have documented caribou
swimming in the ocean (Table 1).

@@

The Fogo Island archipelago is located in Newfoundland, Canada (49°N, 54°W), and
comprises approximately ten islands. While Fogo Island is the largest island
(237.71 km^2^), the archipelago includes at least three larger islands: Western
Indian (77.6 km2), Eastern Indian (38.7 km2) and Change (XX km2) Islands, as
well as numerous smaller islands, including Blundon’s (1.18 km2), North Long
(1.01 km2), South Long (0.48 km2), Kate’s (1.64 km2), and Brother’s (1.59 km2)
Islands (see Figure 1). All islands in the Fogo Island archipelago consist
largely of coniferous and mixed forests of balsam fir (Abies balsamea), black
spruce (Picea mariana), and white birch (Betula payrifera) as well as bogs,
lakes, lichen and rocky barrens.

Forage limitation and over-grazing is a major issue for caribou populations
(Schaefer et al. 2016; Zamin et al. 2017) and could explain movement between
islands. Reduced access to foraging opportunities can result in higher
conspecific competition for resources and lowered reproductive success for adult
females (Schaefer et al. 2016). Forage scarcity has been proposed as a potential
reason why caribou move between arctic islands on the sea-ice in winter (Miller
et al. 1977).  For very small islands, over-grazing by newly arrived caribou may
rapidly limit forage (Bergerud et al. 1990). Without sea ice, movement from the
mainland to an island as well as between islands requires caribou to assess the
trade-off associated with swimming.

@@

Predator avoidance could contribute to swimming behaviour of caribou (Miller
2002). Although caribou co-evolved with wolves (*Canis lupus*) throughout their
range, wolves are extirpated from Newfoundland, including Fogo Island. Coyotes
(*Canis latrans*) and black bears (*Ursus americanus*) are the primary predators
of caribou in Newfoundland (Bastille-Rousseau et al. 2016). While predation by
coyotes on adult female caribou does occur (ref?), it is relatively rare and the
effect of predation on swimming behaviour is more likely the result of
encounters which result in a flight response. For example, during calving,
females may swim to small islands to give birth (Bergerud & Page 1987) or swim
to islands in freshwater lakes with their calves to avoid predation.

@@

Caribou were introduced to Fogo Island in the 1960s and for several decades the
population did not exceed ~100 individuals (Bergerud & Mercer 1989; Newfoundland
and Labrador Wildlife Division, unpublished data). During the 1990s, population
density reached 300 animals and anecdotal evidence suggests that caribou began
to occupy other islands in the archipelago. We surmise that, as density
increased over time, competition among conspecifics and density-dependent
habitat selection resulted in expansion of the population to nearby islands.
Following density-dependent habitat selection theory, caribou should swim to new
islands when the average fitness of individuals on the starting island exceeds
the density-fitness equilibrium (Morris 1987). Animals would therefore be
predicted to swim to new islands (i.e. settle new habitat patches) to reach a
fitness equilibrium. While this is an ultimate explanation to a series of
proximate observations, it is possible for density-dependent habitat selection
to operate at such a fine-scale (Webber & Vander Wal 2018). Evidence from the
Sable Island horse (Equus ferus caballus) system corroborates our suggestion
that swimming among islands could be driven by density-dependent habitat
selection (van Beest et al. 2014). For horses, selection for high quality
habitat was highest when population density was low, but individuals tended to
settle in lower quality habitats as population density increased, following
Ideal Free Distribution theory (van Beest et al. 2014). Taken together, caribou
may follow density-dependent habitat selection (Wittmer et al. 2007) and as
population density increased, forage was depleted and competition increased.

Our observations represent the most frequent and numerous swimming events for
caribou in the ocean (for comparison see Table 1). While past observations (e.g.
Miller 1995, 2002; Jeffery et al. 2007; Ricca et al. 2012) have noted the unique
nature of caribou swimming in the ocean, we suggest that this behaviour is
likely more common than previously thought for caribou living on oceanic
islands. We posit forage limitation and density-dependent habitat selection is
an ultimate explanation for why caribou swimming in the ocean while predator
avoidance is a proximate explanation. Although our inference is limited to
observations, we suggest that because swimming behaviour is so common in our
system, that density-dependent habitat selection and predation are plausible
mechanisms explaining swimming behaviour in Fogo Island caribou.

>>>>

---
title: Interisland movements of woodland caribou in Newfoundland
author: Quinn M.R. Webber^1\*^, Jack G. Hendrix^1^, Alec L. Robitaille^2^,  Eric Vander Wal^1,2^
date: |
  | ^1^Cognitive and Behavioural Ecology Interdisciplinary Program, Memorial University of Newfoundland
  | ^2^Department of Biology, Memorial University of Newfoundland, Canada
  | \*Corresponding author: Quinn M.R. Webber (webber.quinn@gmail.com)
output: 
   # github_document
  # bookdown::pdf_document2:
  #    number_sections: true
  #    toc: false
  bookdown::word_document2:
  #  reference_docx: template.docx
bibliography: ../references/knitcite.bib
#csl: csl.csl
---

@@

## Introduction
Caribou (*Rangifer tarandus*) are exceptional swimmers. Ample evidence exists
that caribou swim in streams, rivers, and lakes during migration (Leblond et al.
2016) to avoid predators [@Bergerud_1985] and access islands during calving
(Bergerud et al. 1990). Swimming for quadrupedal terrestrial mammals can be
thought of as a set of trade-offs. Costs include high energy expenditure and
risk of drowning. Specifically, even though caribou are adept swimmers, the
energetic expenditure associated with swimming for quadrupedal mammals is
significantly higher than walking or running (Fish 1993), whiledrowning
is also possible (Miller & Gunn 1985). Despite a large number of coastal and
island caribou herds, only a handful of examples have documented caribou
swimming in the ocean (Table 1).

@@

<!--I don't know if we need this and if yes, where it will go - thoughts?--> 
The Fogo Island archipelago is located in Newfoundland, Canada (49°N, 54°W), and
comprises approximately ten islands. While Fogo Island is the largest island
(237.71 km^2^), the archipelago includes at least three larger islands: Western
Indian (77.6 km2), Eastern Indian (38.7 km2) and Change (XX km2) Islands, as
well as numerous smaller islands, including Blundon’s (1.18 km2), North Long
(1.01 km2), South Long (0.48 km2), Kate’s (1.64 km2), and Brother’s (1.59 km2)
Islands (see Figure 1). All islands in the Fogo Island archipelago consist
largely of coniferous and mixed forests of balsam fir (Abies balsamea), black
spruce (Picea mariana), and white birch (Betula payrifera) as well as bogs,
lakes, lichen and rocky barrens.

## Forage limitation
Forage limitation and over-grazing is a major issue for caribou populations
(Schaefer et al. 2016; Zamin et al. 2017) and could explain movement between
islands. Reduced access to foraging opportunities can result in higher
conspecific competition for resources and lowered reproductive success for adult
females (Schaefer et al. 2016). Forage scarcity has been proposed as a potential
reason why caribou move between arctic islands on the sea-ice in winter (Miller
et al. 1977).  For very small islands, over-grazing by newly arrived caribou may
rapidly limit forage (Bergerud et al. 1990). Without sea ice, movement from the
mainland to an island as well as between islands requires caribou to assess the
trade-off associated with swimming.

@@

## Predator avoidance
Predator avoidance could contribute to swimming behaviour of caribou (Miller
2002). Although caribou co-evolved with wolves (*Canis lupus*) throughout their
range, wolves are extirpated from Newfoundland, including Fogo Island. Coyotes
(*Canis latrans*) and black bears (*Ursus americanus*) are the primary predators
of caribou in Newfoundland (Bastille-Rousseau et al. 2016). While predation by
coyotes on adult female caribou does occur (ref?), it is relatively rare and the
effect of predation on swimming behaviour is more likely the result of
encounters which result in a flight response. For example, during calving,
females may swim to small islands to give birth (Bergerud & Page 1987) or swim
to islands in freshwater lakes with their calves to avoid predation.

@@

## Summary and conclusion
Caribou were introduced to Fogo Island in the 1960s and for several decades the
population did not exceed ~100 individuals (Bergerud & Mercer 1989; Newfoundland
and Labrador Wildlife Division, unpublished data). During the 1990s, population
density reached 300 animals and anecdotal evidence suggests that caribou began
to occupy other islands in the archipelago. We surmise that, as density
increased over time, competition among conspecifics and density-dependent
habitat selection resulted in expansion of the population to nearby islands.
Following density-dependent habitat selection theory, caribou should swim to new
islands when the average fitness of individuals on the starting island exceeds
the density-fitness equilibrium (Morris 1987). Animals would therefore be
predicted to swim to new islands (i.e. settle new habitat patches) to reach a
fitness equilibrium. While this is an ultimate explanation to a series of
proximate observations, it is possible for density-dependent habitat selection
to operate at such a fine-scale (Webber & Vander Wal 2018). Evidence from the
Sable Island horse (Equus ferus caballus) system corroborates our suggestion
that swimming among islands could be driven by density-dependent habitat
selection (van Beest et al. 2014). For horses, selection for high quality
habitat was highest when population density was low, but individuals tended to
settle in lower quality habitats as population density increased, following
Ideal Free Distribution theory (van Beest et al. 2014). Taken together, caribou
may follow density-dependent habitat selection (Wittmer et al. 2007) and as
population density increased, forage was depleted and competition increased.

Our observations add to the evidence that caribou can, and occaissionaly do,
swim in the ocean (Table 1). While past observations (e.g. Miller 1995, 2002;
Jeffery et al. 2007; Ricca et al. 2012) have noted the unique nature of caribou
swimming in the ocean, we suggest that this behaviour is likely more common than
previously thought for caribou living on oceanic islands. We posit forage
limitation and the associated density-dependent habitat selection is an ultimate
explanation for why caribou swimming in the ocean while predator avoidance is a
proximate explanation. Although our inference is limited to observations, our
observations suggest that islands represent discrete foraging patches for
terrestrial animals that could influence fitness via increased foraging
opportunities and safety from predators.
